Tidy data structure to support exploration and modeling of temporal data

### 2018-12-11 main.Rmd
rewording

---
title: "Tidy data structure to support exploration and modeling of temporal data"
author:
  - familyname: Wang
    othernames: Earo
    address: Department of Econometrics and Business Statistics, \newline Monash University, VIC 3800 \newline Australia.
    email : earo.wang@monash.edu
    correspondingauthor: true
  - familyname: Cook
    othernames: Dianne
    address: Department of Econometrics and Business Statistics, \newline Monash University, VIC 3800 \newline Australia.
    email : dicook@monash.edu
  - familyname: Hyndman
    othernames: Rob J
    address: Department of Econometrics and Business Statistics, \newline Monash University, VIC 3800 \newline Australia.
    email : rob.hyndman@monash.edu
abstract: 'Mining temporal data for information is often inhibited by a multitude of time formats: irregular or multiple time intervals, multiple observational units or repeated measurements on multiple individuals, heterogeneous data types. Time series models, in particular, the software supporting time series forecasting makes strict assumptions on data to be provided, typically a matrix of numeric data with an implicit time index. Going from raw data to model-ready data is painful. This work presents a cohesive and conceptual framework for organizing and manipulating temporal data, which in turn flows into visualization and forecasting routines. Tidy data principles are applied, and extended to temporal data: (1) mapping the semantics of a dataset into its physical layout, (2) including an explicitly declared index variable representing time, (3) incorporating a "key" comprised of single or multiple variables to uniquely identify units over time. This tidy data representation most naturally supports thinking of operations on the data as building blocks, forming part of a "data pipeline" in time-based context. A sound data pipeline facilitates a fluent and transparent workflow for analyzing temporal data. Applications are included to illustrate tidy temporal data structure, data pipeline structure and usage. The infrastructure of tidy temporal data has been implemented in the R package **tsibble**.'
keywords: "temporal data, time series, data structures, data wrangling, tidy data, R, forecasting, data science, exploratory data analysis"
wpnumber: no/yr
jelcodes: C10,C14,C22
blind: false
cover: true
toc: false
bibliography: [references.bib, rpkgs.bib]
biblio-style: authoryear-comp
output:
  MonashEBSTemplates::workingpaper:
    fig_caption: yes
    fig_height: 5
    fig_width: 8
    includes:
      in_header: preamble.tex
    keep_tex: no
    number_sections: yes
    citation_package: biblatex
---

@@

@wickham2014tidy coined the term "tidy data", which is a rephrasing of the second and third normal forms in relational databases but in a way that makes more sense to data scientists by referring rows to observations and columns to variables. The principles of "tidy data" attempt to standardize the mapping of the semantics of a dataset to its physical representation. This data structure is the fundamental unit of the **tidyverse**, which is a collection of R packages designed for data science; the prevalence of **tidyverse** reversely testifies to the novelty of "tidy data". The **tidyverse** provides abstract yet functional grammars to manipulate and visualize data in easier-to-comprehend manners. One of the **tidyverse** packages, **dplyr** [@R-dplyr] showcases the value of a grammar as a principled vehicle to transform data for a wide range of data challenges, providing a consistent set of verbs: `mutate()`, `select()`, `filter()`, `summarise()`, and `arrange()`. Each verb focuses on a singular task. Almost all data tasks can be rephrased and tackled with these five key verbs, by composing them sequentially.

Community effort for building tidy ecosystem

@@

The choice of tidy representation of temporal data is made from a data-centric perspective, which is taken in the light of the operations that are to be performed on the data. Contextual semantics are introduced to tidy data in order to support more intuitive time-related manipulations and enlighten new perspectives for time series model inputs. Index, key and time interval are the three stone pillars to this new semantically-structured temporal data, which we name as "tsibble".

@@

The tidy data abstraction lays a pipeline infrastructure for data analysis modules, transformation, visualization and modelling. Each module communicates between each other, requiring tidy input, producing tidy output, and consequently chains a series of operations together to accomplish the analytic tasks at hand.

@@

The time-based pipeline shepards raw temporal data through to time series analysis, and plots. It is advised to scrutinize identical entries of key and index in the right beginning. Duplicates signal the data quality issue, which would likely affect succeeding analyses and hence decision making. Analysts are encouraged to gaze at data as earlier as possible and reason about the process of data cleaning. When the data meets the tsibble standard, it flows into the exploration stage.

Time series models typically assume that the input is a complete and regularly-spaced series, which most temporal data can barely satisfy. Temporal data becomes more granular in time resolution and more disaggregated to individual level. This inevitably goes with implicit missing values and noisiness. Data needs transformation to some extent for modelling. A suite of verbs are introduced to flatten the lumpy path from temporal data to the object that directs to modelling under the tsibble framework, as well as to facilitate transforming tsibble in various kinds of shapes for analysis and visualization. The principle that underpins most verbs is a tsibble in and a tsibble out, thereby striving to retain a valid tsibble by automating index and key updates under the hood. If a tsibble cannot hold, for example swiping index off, an error prompts users to complain and suggest alternatives. This warrants users least surprise and reminds them of time awareness.

Transformation assembles a set of modules. A unit that makes up a module is function or preferably "verb". A tsibble is an object, conceptually considered as a noun, and hence an action performed on the object can be phrased as a verb. Each verb focuses on one thing and achieve its goal. The verb should be self-explanatory to advise what it is supposed to do or fail, for example `filter()` picking observations, and `select()` picking variables. These general-purpose verbs are made available in the **tidyverse** suite. When manipulating in temporal context, these verbs are adapted to time domain. A perceivable difference is summarizing variables between data frame and tsibble. The former will reduce to a single summary, whereas the latter will obtain the index and their corresponding summaries. New tsibble-specific verbs are proposed to expand the **tidyverse** vocabulary. We believe that users, who are already familiar with **tidyverse**, will experience a gentle learning curve for mastering tsibble verbs and glide into temporal data analysis with low cognitive load.

\begin{table}[!hbt]
\caption{A test}\label{tab:verb-table}
\centering
\begin{tabular}{ lll }
\toprule
 & Verb & Description \\ \midrule
\multirow{4}{*}{Time gaps} & \texttt{has\_gaps()} & Test if a tsibble has gaps in time \\
 & \texttt{scan\_gaps()} & Report implicit missing entries \\
 & \texttt{count\_gaps()} & Summarize time gaps \\
 & \texttt{fill\_gaps()} & Fill in gaps by values and functions \\ \hline
\multirow{4}{*}{Row-wise} & \textbf{\texttt{filter()}} & Pick rows based on conditions \\
 & \texttt{filter\_index()} & Provide a shorthand for time subsetting \\
 & \textbf{\texttt{slice()}} & Select rows based on row positions \\
 & \textbf{\texttt{arrange()}} & Sort the ordering of row by variables \\ \hline
\multirow{4}{*}{Column-wise} & \textbf{\texttt{select()}} & Pick columns by variables \\
 & \textbf{\texttt{mutate()}} & Add new variables \\
 & \textbf{\texttt{transmute()}} & Add new variables \\
 & \textbf{\texttt{summarise()}} & Aggregate values over time \\ \hline
\multirow{3}{*}{Group-wise} & \texttt{index\_by()} & Group by new time index \\
 & \textbf{\texttt{group\_by()}} & Group by one or more variables \\
 & \texttt{group\_by\_key()} & Group by key variables \\ \hline
\multirow{4}{*}{Reshape data} & \textbf{\texttt{gather()}} & David Batty \\
 & \textbf{\texttt{spread()}} & Eirik Bakke \\
 & \textbf{\texttt{nest()}} & Eirik Bakke \\
 & \textbf{\texttt{unnest()}} & Jody Morris \\
\bottomrule
\end{tabular}
\end{table}

@@

* **rolling window**: `slide()`, `tile()`, `stretch()`

>>>>

---
title: "Tidy data structure to support exploration and modeling of temporal data"
author:
  - familyname: Wang
    othernames: Earo
    address: Department of Econometrics and Business Statistics, \newline Monash University, VIC 3800 \newline Australia.
    email : earo.wang@monash.edu
    correspondingauthor: true
  - familyname: Cook
    othernames: Dianne
    address: Department of Econometrics and Business Statistics, \newline Monash University, VIC 3800 \newline Australia.
    email : dicook@monash.edu
  - familyname: Hyndman
    othernames: Rob J
    address: Department of Econometrics and Business Statistics, \newline Monash University, VIC 3800 \newline Australia.
    email : rob.hyndman@monash.edu
abstract: 'Mining temporal data for information is often inhibited by a multitude of formats: irregular or multiple time intervals, point event that needs aggregating, multiple observational units or repeated measurements on multiple individuals, heterogeneous data types. Time series models, in particular, the software supporting time series forecasting makes strict assumptions on the data to be provided, typically a matrix of numeric data with implicit time indices. Going from raw data to model-ready data is painful. This work presents a cohesive and conceptual framework for organizing and manipulating temporal data, which in turn flows into visualization and forecasting routines. Tidy data principles are extended to temporal data: (1) mapping the semantics of a dataset into its physical layout, (2) including an explicitly declared index variable representing time, (3) incorporating a "key" comprised of single or multiple variables to uniquely identify units over time. This tidy data representation most naturally supports thinking of operations on the data as building blocks, forming part of a "data pipeline" in time-based context. A sound data pipeline facilitates a fluent and fluid workflow for analyzing temporal data. The infrastructure of tidy temporal data has been implemented in the R package **tsibble**.'
keywords: "temporal data, time series, data structures, data wrangling, tidy data, R, forecasting, data science, exploratory data analysis"
wpnumber: no/yr
jelcodes: C10,C14,C22
blind: false
cover: true
toc: false
bibliography: [references.bib, rpkgs.bib]
biblio-style: authoryear-comp
output:
  MonashEBSTemplates::workingpaper:
    fig_caption: yes
    fig_height: 5
    fig_width: 8
    includes:
      in_header: preamble.tex
    keep_tex: no
    number_sections: yes
    citation_package: biblatex
---

@@

@wickham2014tidy coined the term "tidy data", which is a rephrasing of the second and third normal forms in relational databases but in a way that makes more sense to data scientists by referring rows to observations and columns to variables. The principles of "tidy data" attempt to standardize the mapping of the semantics of a dataset to its physical representation. This data structure is the fundamental unit of the **tidyverse**, which is a collection of R packages designed for data science; the prevalence of **tidyverse** reversely testifies to the novelty of "tidy data". The **tidyverse** provides abstract yet functional grammars to manipulate and visualize data in easier-to-comprehend manners. One of the **tidyverse** packages, **dplyr** [@R-dplyr] showcases the value of a grammar as a principled vehicle to transform data for a wide range of data challenges, providing a consistent set of verbs: `mutate()`, `select()`, `filter()`, `summarise()`, and `arrange()`. Each verb focuses on a singular task. Most common data tasks can be rephrased and tackled with these five key verbs, by composing them sequentially.

The **tidyverse** way of thinking largely formalizes exploratory data analysis. Tidy concepts have been extended to other domains in R community.

@@

The choice of tidy representation of temporal data is made from a data-centric perspective, which is taken in the light of the operations that are to be performed on the data. Contextual semantics are introduced to tidy data in order to support more intuitive time-related manipulations and enlighten new perspectives for time series model inputs. Index, key and time interval are the three stone pillars to this new semantically-structured temporal data, which we name as "tsibble", inspired by "ts" and "tibble".

@@

The tidy data abstraction lays a pipeline infrastructure for data analysis modules of transformation, visualization and modelling. Each module communicates between each other, requiring tidy input, producing tidy output, chaining a series of operations together to accomplish the analytic tasks.

@@

The time-based pipeline shepards raw temporal data through to time series analysis, and plots. It is possible and recommended to check for identical entries of key and index before analysis. Duplicates signal the data quality issue, which would likely affect succeeding analyses and hence decision making. Analysts are encouraged to gaze at data early and reason about the process of data cleaning. When the data meets the tsibble standard, it flows neatly into the exploration stage.

Time series models typically assume that the input is a complete and regularly-spaced series, which most temporal data can barely satisfy. Temporal data becomes more granular in time resolution and more disaggregated to individual level. This inevitably goes with implicit missing values and noisiness. Data needs transformation to some extent for modelling. A suite of verbs are introduced to flatten the lumpy path from temporal data to an object that can be directly modeled in the tsibble framework, as well as transformed for visualization. The principle that underpins most verbs is a tsibble in and a tsibble out, thereby striving to retain a valid tsibble by automating index and key updates under the hood. Attention has been paid to error handling. If a tsibble cannot be maintained in the output of a pipeline module, for example the index is dropped by aggregation, an error informs users of the problem and suggests alternatives. This avoids negatively surprising users and reminds them of time context.

Transformation assembles a set of modules. A unit that makes up a module is function or preferably "verb". A tsibble is an object, conceptually considered as a noun, and hence an action performed on the object can be phrased as a verb. Each verb focuses on one thing and achieve its goal. The verb should be self-explanatory to advise what it is supposed to do or fail, for example `filter()` picks observations, `select()` picks variables, and `left_join()` joins two tables. These general-purpose verbs are made available in the **tidyverse** suite. When manipulating in temporal context, these verbs are adapted to time domain. A perceivable difference is summarizing variables between data frame and tsibble. The former will reduce to a single summary, whereas the latter will obtain the index and their corresponding summaries. New tsibble-specific verbs are proposed to expand the **tidyverse** vocabulary. We believe that users, who are already familiar with **tidyverse**, will experience a gentle learning curve for mastering tsibble verbs and glide into temporal data analysis with low cognitive load.

\begin{table}[!hbt]
\caption{A test}\label{tab:verb-table}
\centering
\begin{tabular}{ lll }
\toprule
 & Verb & Description \\ \midrule
\multirow{4}{*}{Time gaps} & \texttt{has\_gaps()} & Test if a tsibble has gaps in time \\
 & \texttt{scan\_gaps()} & Report implicit missing entries \\
 & \texttt{count\_gaps()} & Summarize time gaps \\
 & \texttt{fill\_gaps()} & Fill in gaps by values and functions \\ \hline
\multirow{4}{*}{Row-wise} & \textbf{\texttt{filter()}} & Pick rows based on conditions \\
 & \texttt{filter\_index()} & Provide a shorthand for time subsetting \\
 & \textbf{\texttt{slice()}} & Select rows based on row positions \\
 & \textbf{\texttt{arrange()}} & Sort the ordering of row by variables \\ \hline
\multirow{4}{*}{Column-wise} & \textbf{\texttt{select()}} & Pick columns by variables \\
 & \textbf{\texttt{mutate()}} & Add new variables \\
 & \textbf{\texttt{transmute()}} & Add new variables \\
 & \textbf{\texttt{summarise()}} & Aggregate values over time \\ \hline
\multirow{3}{*}{Group-wise} & \texttt{index\_by()} & Group by new time index \\
 & \textbf{\texttt{group\_by()}} & Group by one or more variables \\
 & \texttt{group\_by\_key()} & Group by key variables \\ \hline
\multirow{4}{*}{Reshape data} & \textbf{\texttt{gather()}} & David Batty \\
 & \textbf{\texttt{spread()}} & Eirik Bakke \\
 & \textbf{\texttt{nest()}} & Eirik Bakke \\
 & \textbf{\texttt{unnest()}} & Jody Morris \\ \hline
\multirow{6}{*}{Join data} & \textbf{\texttt{left\_join()}} & David Batty \\
 & \textbf{\texttt{right\_join()}} & Eirik Bakke \\
 & \textbf{\texttt{full\_join()}} & Eirik Bakke \\
 & \textbf{\texttt{inner\_join()}} & Jody Morris \\
 & \textbf{\texttt{semi\_join()}} & Jody Morris \\
 & \textbf{\texttt{anti\_join()}} & Jody Morris \\
\bottomrule
\end{tabular}
\end{table}

@@

A set of verbs have been implemented to conduct temporal operations. There are three kinds of sliding window operations:

* `slide()`/`slide2()`/`pslide()`: sliding window with overlapping observations.
* `tile()`/`tile2()`/`ptile()`: tiling window without overlapping observations.
* `stretch()`/`stretch2()`/`pstretch()`: fixing an initial window and expanding to include more observations.

As the name indicates, `slide()` expects one input, `slide2()` two inputs, and `pslide()` multiple inputs. This family of window functions features functional programming. Any function can be mapped and rolled over a block of elements given a specified window size. For type stability, the functions always return lists. Other variants including `slide_lgl()`, `slide_int()`, `slide_dbl()`, `slide_chr()` return vectors of the corresponding type, as well as `slide_dfr()` and `slide_dfc()` for row-binding and column-binding data frames respectively. Their multiprocessing equivalents prefixed by `future_` enable rolling in parallel. This family of functions empowers users to incorporate window-related operations in their workflows.
